---
title: "Untitled"
output: html_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Remove doublets and score cell cyle
Continue with the rds of clusterByLeiden0.2.rds

```{r}
library(Seurat)
library(DoubletFinder)
library(here)
library(ggplot2)
library(dplyr)
```

```{r}
merged <- readRDS(here("results","clusterByLeiden0.2.rds"))
print(merged)
```

# make sure the default assay is RNA and DefaultLayer is data
```{r}
DefaultAssay(merged) <- "RNA"
DefaultLayer(merged[["RNA"]]) <- "data"
```


# check the cell doublets and score cell cycle

The expected doublet rate for 10x Genomics data is strongly dependent on the number of cells loaded during GEM (Gel Bead-in-Emulsion) generation, and 10x has published fairly detailed guidance.

10K cells per sample Expected Doublet Rate is ~8.0%
Or use a slightly conservative estimate like 5–6% globally
since my dataset has been filtered out top and bottom 5% which removed many doublets,
I also have explored using 8%, now explore with 5%

Term	Meaning
pK	Number of neighbors to consider for doublet scoring
BCmetric	Score of how well artificial vs. real cells are separated
pANN	% of nearest neighbors that are artificial (per cell)
pN	% of artificial doublets added to the dataset

```{r}
# Run DoubletFinder
sweep.res <- paramSweep(merged, PCs = 1:30, sct = FALSE)
sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
bcmvn <- find.pK(sweep.stats)
```


```{r}
# Convert pK column to numeric (if needed)
bcmvn$pK <- as.numeric(as.character(bcmvn$pK))

# Find pK with highest BCmetric
best_index <- which.max(bcmvn$BCmetric)
best_pK <- bcmvn$pK[best_index]
best_metric <- bcmvn$BCmetric[best_index]

cat("Best pK:", best_pK, "with BCmetric:", best_metric, "\n")

# Plot with labeled axes and a line at best pK
ggplot(bcmvn, aes(x = pK, y = BCmetric)) +
  geom_point(color = "dodgerblue") +
  geom_line(color = "dodgerblue") +
  geom_vline(xintercept = best_pK, color = "red", linetype = "dashed") +
  annotate("text", x = best_pK, y = best_metric + 0.01, label = paste0("Best pK = ", best_pK),
           hjust = 0, color = "red", size = 3) +
  labs(
    title = "BCmetric vs pK",
    x = "pK (neighborhood size)",
    y = "BCmetric (classification performance)"
  ) +
  theme_minimal()
```

pN: This is almost always left at 0.25, the recommended default: 0.25
pK: choose according to the above BCmetric plot

Check when using 5%
```{r}

# Estimate expected doublets: ~1% for 10x, or use nExp = 0.05 * ncol(merged)
nExp_poi <- round(0.05 * ncol(merged))  # 5% doublets

merged <- doubletFinder(
  merged,
  PCs = 1:30,
  pN = 0.25,
  pK = best_pK,        # ← use your top pK here
  nExp = nExp_poi, # ← use your top nExp_poi here
  sct = FALSE
)

table(merged@meta.data[[grep("DF.classification", names(merged@meta.data), value = TRUE)]])
DimPlot(merged, group.by = grep("DF.classification", names(merged@meta.data), value = TRUE))

```
```{r}
# Inspect results
df_col <- grep("DF.classifications", names(merged@meta.data), value = TRUE)[1]
table(merged@meta.data[[df_col]])
DimPlot(merged, group.by = df_col)
```


```{r}
nExp_poi <- round(0.08 * ncol(merged))  # 5% doublets

merged <- doubletFinder(
  merged,
  PCs = 1:30,
  pN = 0.25,
  pK = best_pK,        # ← use your top pK here
  nExp = nExp_poi, # ← use your top nExp_poi here
  sct = FALSE
)

```

```{r}
colnames(merged@meta.data)
```

```{r}
# Inspect results
df_col <- grep("DF.classifications", names(merged@meta.data), value = TRUE)[3]
table(merged@meta.data[[df_col]])
DimPlot(merged, group.by = df_col)
```

# remove doublets using the ratio 5%

First remove unnecessary meta data saved in the objects: pANN...DF...

```{r}
print(names(merged@meta.data))
```


```{r}
cols_to_remove <- c(
  "pANN_0.25_0.24_1513",
  "pANN_0.25_0.25_2420",
  "DF.classifications_0.25_0.24_1513",
  "DF.classifications_0.25_0.25_2420"
)

merged@meta.data <- merged@meta.data[, !(colnames(merged@meta.data) %in% cols_to_remove)]

```

"pANN_0.25_0.25_1513": 0.25 means generate %25 number of the total cells as artificial doulbets, secnodn 0.25 means %25 real cells were used for generate ANN score, 1513 means there are 1513 doublet cells
```{r}
print(names(merged@meta.data))
```

```{r}
# Step 1: Identify singlet cells using your metadata column
df_col <- grep("DF.classifications", names(merged@meta.data), value = TRUE)[1]
singlet_cells <- colnames(merged)[merged@meta.data[[df_col]] == "Singlet"]

# Step 2: Subset using those cell names
merged_noDoublet <- subset(merged, cells = singlet_cells)
```


```{r}
print(merged_noDoublet)
DimPlot(merged_noDoublet)

```
Removing doublets can slightly shift the structure of your data. It's a good idea to re-run the belows
I didn't use leiden anymore, since I didn't notice apprarent difference in clusters and it use too much resouces, takes longer, need python env

```{r}
library(magrittr)  # enables the pipe (%>%)

merged_noDoublet <- merged_noDoublet %>%
  ScaleData() %>%
  RunPCA(npcs = 30) %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters(resolution = 0.2) %>%
  RunUMAP(dims = 1:30)
```

save a temp rds
```{r}
#saveRDS(merged_noDoublet, file = here("results", "noDoublets.rds"))
```

# Score Cell Cycle in Seurat

Seurat includes canonical S phase and G2/M phase genes:

```{r}
s.genes <- Seurat::cc.genes.updated.2019$s.genes
g2m.genes <- Seurat::cc.genes.updated.2019$g2m.genes

```

Make sure your data is normalized first 

- Adds three new columns to merged@meta.data:
-- "S.Score" — how strongly each cell expresses S phase genes
-- "G2M.Score" — expression of G2/M genes
-- "Phase" — inferred cell cycle phase: "G1", "S", or "G2M"
- Optionally sets the Seurat identity class (Idents) to "Phase" if set.ident = TRUE

```{r}
merged_noDoublet <- CellCycleScoring(
  merged_noDoublet,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = TRUE
)

```
```{r}
DimPlot(merged_noDoublet, group.by = "seurat_clusters")
```

Visualize cell cycle
```{r}
DimPlot(merged_noDoublet, group.by = "Phase", split.by = "condition", label = TRUE)
```
```{r}
DimPlot(merged_noDoublet, group.by = "Phase", label = TRUE)
```

```{r}
g1_cells <- colnames(merged_noDoublet)[merged_noDoublet$Phase == "G1"]
merged_G1 <- subset(merged_noDoublet, cells = g1_cells)
DimPlot(merged_G1, group.by = "Phase", split.by = "condition", label = TRUE)
```


```{r}
g2_cells <- colnames(merged_noDoublet)[merged_noDoublet$Phase == "G2M"]
merged_G2M <- subset(merged_noDoublet, cells = g2_cells)
DimPlot(merged_G2M, group.by = "Phase", split.by = "condition", label = TRUE)
```

```{r}
S_cells <- colnames(merged_noDoublet)[merged_noDoublet$Phase == "S"]
merged_S <- subset(merged_noDoublet, cells = S_cells)
DimPlot(merged_S, group.by = "Phase", split.by = "condition", label = TRUE)
```


```{r}
VlnPlot(merged_noDoublet, features = c("S.Score", "G2M.Score"), group.by = "Phase")
```
save a temp rds
```{r}
saveRDS(merged_noDoublet, file = here("results", "noDoublets_scoreCellCycle.rds"))
```

