---
title: "Untitled"
output: html_document
date: "2025-03-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

sudo apt-get update && sudo apt-get install -y   libcurl4-openssl-dev   libssl-dev   libxml2-dev   libharfbuzz-dev   libfribidi-dev   libpng-dev   libcairo2-dev   libfontconfig1-dev   libtiff5-dev   libjpeg-dev   libgl1-mesa-dev   libglu1-mesa-dev   libgdal-dev

```{r}
install.packages(c(
  "Seurat",
  "patchwork",
  "sctransform",
  "ggplot2",
  "dplyr",
  "reticulate",
  "tidyverse"
))

```

```{r}
install.packages("BiocManager")
BiocManager::install(c("DESeq2", "hdf5r"))

```

```{r}
#install.packages("SeuratData") #notworking
install.packages("remotes")
remotes::install_github("satijalab/seurat-data")
library(SeuratData)
SeuratData::InstallData("pbmc3k")

```

# install the dependency of Azimuth
sudo apt-get update && sudo apt-get install -y libgsl-dev
```{r}
# dpendence for TFBSTools
BiocManager::install("DirichletMultinomial", update = FALSE, ask = FALSE)
```

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "BSgenome.Hsapiens.UCSC.hg38",
  "EnsDb.Hsapiens.v86",
  "TFBSTools"
))

```
```{r}
BiocManager::install("HDF5Array")
install.packages("assertthat")
```

```{r}
#remotes::install_github("const-ae/glmGamPoi", verbose = TRUE) # failed
BiocManager::install("glmGamPoi")

```

```{r}
remotes::install_github("satijalab/azimuth")
```

```{r}
#BiocManager::install("BPCells") # failed

```

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(hdf5r)
library(dplyr)
library(reticulate)
library(DESeq2) # if want to perform DE using this option
library(reticulate)
#library(BPCells)
library(tidyverse)
```

Use this to snapshot your working packages:
renv maintains a local cache of data on the filesystem, located at:
  - "~/.cache/R/renv"

o Recreate This Setup on Another VM Later
Copy your project folder (with renv.lock and renv/) and run:

```{r}
install.packages("renv")
renv::init()
```


That ensures you never go through this dependency maze again 😄
```{r}
renv::status()
```
```{r}
library(renv)
renv::settings$ignored.packages("BPCells")
renv::settings("ignored.files", value = c("notebooks"))
renv::snapshot()
```

