---
title: "annotation using customized reference dataset"
output: html_document
date: "2023-08-02"
---

https://github.com/satijalab/azimuth/issues/21

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
```
```{r}
library(Seurat)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(patchwork)
```

# read the reference dataset containing cell type info

This was downloaded from the group in Germany in the paper of Temperail sars transcriptomics in Nature Communications
https://www.nature.com/articles/s41467-021-25030-7

It has both curated annotation as "celltype" as well as annoation using tabula_muris as "predited.id.tabula_muris"
Apparently, the celltypes from these two methods are quit different

```{r}
ref <- readRDS("ma_int.rds")
ref
```

```{r}
colnames(ref@meta.data)
```

```{r}
head(ref[["celltype"]], 15)
```

```{r}
range(ref[["SCoV2_load"]])
```

```{r}
range(ref[["SCoV2_sum"]])
```


```{r}
head(ref[["prev.cluster"]])
```
```{r}
head(ref[["predicted.id.tabula_muris"]], 15)
```

# generate hamster ref
# first downsample to 2000 
```{r}
ref <- subset(x = ref, downsample = 2000)
ref
```

# can refer to the script but need some troubleshooting
https://github.com/satijalab/azimuth-references/blob/master/human_motorcortex/scripts/export.R
# fix errors: change meta column name, run SCTransform

```{r}
ref$cell.type.tabula_muris <- factor(ref$predicted.id.tabula_muris)
```

```{r}
ref <- SCTransform(ref)
ref <- RunUMAP(object = ref, reduction = "pca", dims = 1:30, return.model = TRUE)
```

```{r}
ref <- AzimuthReference(
  object = ref,
  refUMAP = "umap",
  refDR = "pca",
  refAssay = "SCT",
  metadata = c("celltype", "seurat_cluster", "cell.type.tabula_muris"), # by mistakes omitting s in "seurat_clusters"
  dims = 1:50,
  k.param = 31,
  reference.version = "1.0.0"
)
```


```{r}
ref
```

# must need to save them into a folder, whose name will be used as the reference name of in RunAzimuth
2 files must be inside this folder: ref.Rds and idx,annoy

```{r}
SaveAnnoyIndex(object = ref[["refdr.annoy.neighbors"]], file = file.path("hamsterLungRef/idx.annoy"))
saveRDS(object = ref, file = file.path("hamsterLungRef/ref.Rds"))
```


```{r}
colnames(ref@meta.data)
```

# read my own dataset
```{r}
epsilon <- readRDS("epsilon_sct.rds")
"M" %in% rownames(epsilon)
```

# lost my Sars Genes if using default lungref ref datasets
strange thing here: seems to only human gene name
# annotation with default lung ref
```{r}
epsilon_az <- RunAzimuth(epsilon, "lungref")
"M" %in% rownames(epsilon_az) # lost Sars genes at these step
```

# annotate using custmized ref
```{r}
epsilon_hamster <- RunAzimuth(epsilon, "hamsterLungRef")
"M" %in% rownames(epsilon_hamster)
```


```{r}
"M" %in% rownames(epsilon_hamster)
```

# save
```{r}
saveRDS(epsilon_hamster, file = "epsilon_sct_az_hst.rds")
```


```{r}
names(epsilon_hamster)
```
```{r}
colnames(epsilon_hamster@meta.data)
```

# plot sars and cell types
```{r}
sars_genes <- c("S", "M", "N", "ORF1ab", "ORF8")
p1 <- FeaturePlot(epsilon_hamster, features = sars_genes, ncol = 3)
ggsave("sars_umap.pdf", p1, width = 8, height = 6)

p <- DimPlot(epsilon_hamster, group.by = "seurat_clusters")
ggsave("clusters.pdf", p)

p <- DimPlot(epsilon_hamster, group.by = "predicted.celltype")
ggsave("annotation_hamster.pdf", p)
```

```{r}
p <- DimPlot(epsilon_hamster, group.by = "predicted.cell.type.tabula_muris")
ggsave("new_annotations_muris.pdf", p, width = 10, height = 6)
```

```{r}
p3 <- DimPlot(epsilon_hamster, group.by = "predicted.cell.type.tabula_muris", split.by = "orig.ident")
p4 <- DimPlot(epsilon_hamster, group.by = "predicted.celltype", split.by = "orig.ident")
a_plot <- p3 + p4 + plot_layout(nrow = 2)
ggsave("annotations_split_days.pdf", a_plot, width = 16, height = 12)
```
