---
title: "examine the script of integrate_virus.Rmd which executed on 20230912"
output: html_document
date: "2024-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# contiue using the template of integrate_virus.Rmd on 20230911.

# Previous experiences:
# The integrated obj takes 20G mem, and sinteractive --mem=100g is still challenge; so increase to 240g
# increse the max mem in .Renviron up to 200g

sinteractive --mem=240g --cpus-per-task=12 --gres=lsratch:100
cd $IRF
source myconda
conda active r4seurat
module load rstudio R
rstudio&

setwd("/vf/users/pengl7/IRF/seurat/flex2023/")

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(hdf5r)
library(dplyr)
library(tidyr)
library(DESeq2) # if want to perform DE using this option
library(reticulate)
use_condaenv("/data/pengl7/conda/envs/r4seurat")
py_config()
# need to first create a conda env as r4seurat; conda active r4seurat; conda install leidenalg; conda install numpy pandas; start rstudio&
leidenalg <- import("leidenalg")
```
# get the integrated allvirus rds
# it took 4 mins to read input; note that R may be stucked during saving the allrivus.rds
```{r}
run.name <- "all_virus"
sub.obj <- readRDS("../20230831-integrate/rds/infected_samples/allvirus2_az_splitCtrl.rds")
print(sub.obj)
```

# 20230912 analyze days for each strain
```{r}
obj <- readRDS("allvirus2_az.rds")
```

##################################################################
# this is an example to subset a data for a single strain, 
1. then peform AVG expression by conditions such as days, sars status
2. also count cell numbers by conditions

```{r}
Idents(obj) <- "strains.fine.level"
sub.obj <- subset(obj, idents = "Delta")
run.name <- "Delta"
dir.create(run.name)
# 1. perform AVG expression
# first define active identity
Idents(sub.obj) <- sub.obj$predicted.celltype
# by days
avg.exp <- as.data.frame(AverageExpression(sub.obj, assays = "RNA", group.by = c("days")))
write.csv(avg.exp, file = paste0(run.name, "/", run.name, "_gene_avg_exp_by_days.csv"))
# by sars infection
avg.exp <- as.data.frame(AverageExpression(sub.obj, assays = "RNA", group.by = c("sars.pos")))
write.csv(avg.exp, file = paste0(run.name, "/", run.name, "_gene_avg_exp_by_sars.csv"))

# 2. calculate cell number per type
a.table <- table(sub.obj$predicted.celltype, sub.obj$days)
df <- data.frame(a.table)
df <- pivot_wider(df, id_cols = Var1, names_from = Var2, values_from = Freq)
write.csv(df, file = paste0(run.name, "/", run.name, "_cell_numbers_per_type.csv"), row.names = FALSE)
```

3.also perform DE between groups, here between days

```{r}
# if just comare two groups for one condition
compare <- "D3vsD7andD10"
markers <- FindMarkers(sub.obj, ident.1 = "D03")
write.csv(markers, file = paste0(run.name, "/", run.name,"_", compare,"_DE_genes.csv"))
  
top50 <- markers %>%
  slice_max(n = 50, order_by = avg_log2FC)
write.csv(top50, file = paste0(run.name, "/", run.name, "_", compare,"_top20_genes.csv"))
  
bottom50 <- markers %>%
  slice_min(n = 50, order_by = avg_log2FC)
write.csv(bottom50, file = paste0(run.name, "/", run.name,"_", compare,"_bottom20_genes.csv"))
```

```{r}
# if comare multiple groups for one condition
days <- c("D03", "D07", "D10")
Idents(sub.obj) <- sub.obj$days
# Assuming you have a list of 7 objects named 'compare.list'
compare.list <- days
num_objects <- length(compare.list)

# Loop through each pair of objects and apply your comparison function
for (i in 1:(num_objects - 1)) {
  for (j in (i + 1):num_objects) {
    # Skip self-comparisons
    if (i != j) {
      compare <- paste0(compare.list[[i]], "vs", compare.list[[j]])
      ident1 = compare.list[[i]]
      ident2 = compare.list[[j]]
      markers <- FindMarkers(sub.obj, ident.1 = ident1, ident.2 = ident2)
      write.csv(markers, file = paste0(run.name, "/", run.name,"_", compare,"_marker_genes.csv"))
  
      top50 <- markers %>%
        slice_max(n = 50, order_by = avg_log2FC)
      write.csv(top50, file = paste0(run.name, "/", run.name, "_", compare,"_top50_genes.csv"))
      
      bottom50 <- markers %>%
        slice_min(n = 50, order_by = avg_log2FC)
      write.csv(bottom50, file = paste0(run.name, "/", run.name,"_", compare,"_bottom50_genes.csv"))
    }
  }
}

```

#######################################################
The above is for finding AVG expression, count cell count, and DE on 1 conditons; if want to execute AVG expresion, cell count, and DE on different conditions such as different starin 
using below scripts
#######################################################

```{r}
#strains <- levels(factor(obj$strains.fine.level)) 
strains <- c("Omicron_BA.2","WA01") 

# loop each strain

for (strain in strains){
Idents(obj) <- "strains.fine.level"
sub.obj <- subset(obj, idents = strain)
run.name <- strain
dir.create(run.name)
#perform AVG expression using SCT data
Idents(sub.obj) <- sub.obj$predicted.celltype
avg.exp <- as.data.frame(AverageExpression(sub.obj, assays = "RNA", group.by = c("days")))
write.csv(avg.exp, file = paste0(run.name, "/", run.name, "_gene_avg_exp_by_days.csv"))
avg.exp <- as.data.frame(AverageExpression(sub.obj, assays = "RNA", group.by = c("sars.pos")))
write.csv(avg.exp, file = paste0(run.name, "/", run.name, "_gene_avg_exp_by_sars.csv"))

# calculate cell number per type
a.table <- table(sub.obj$predicted.celltype, sub.obj$days)
df <- data.frame(a.table)
df <- pivot_wider(df, id_cols = Var1, names_from = Var2, values_from = Freq)
write.csv(df, file = paste0(run.name, "/", run.name, "_cell_numbers_per_type.csv"), row.names = FALSE)

# plot
DefaultAssay(sub.obj) <- "integrated"
p2 <- DimPlot(sub.obj, group.by = "predicted.celltype", split.by = "days")
ggsave(filename = paste0(run.name, "/", run.name, ".pdf"), width = 16)

# DE genes: 
# using for loop to compare D3vsD7, D3vsD10, D7vsD10

days <- c("D03", "D07", "D10")
Idents(sub.obj) <- sub.obj$days
# Assuming you have a list of 7 objects named 'compare.list'
compare.list <- days
num_objects <- length(compare.list)

for (i in 1:(num_objects - 1)) {
  for (j in (i + 1):num_objects) {
    # Skip self-comparisons
    if (i != j) {
      compare <- paste0(compare.list[[i]], "vs", compare.list[[j]])
      ident1 = compare.list[[i]]
      ident2 = compare.list[[j]]
      markers <- FindMarkers(sub.obj, ident.1 = ident1, ident.2 = ident2)
      write.csv(markers, file = paste0(run.name, "/", run.name,"_", compare,"_DE_genes.csv"))
  
      top50 <- markers %>%
        slice_max(n = 50, order_by = avg_log2FC)
      write.csv(top50, file = paste0(run.name, "/", run.name, "_", compare,"_top50_genes.csv"))
      
      bottom50 <- markers %>%
        slice_min(n = 50, order_by = avg_log2FC)
      write.csv(bottom50, file = paste0(run.name, "/", run.name,"_", compare,"_bottom50_genes.csv"))
    }
  }
}

# compare D3 to the left (D7+D10)
compare <- "D3vsD7andD10"
markers <- FindMarkers(sub.obj, ident.1 = "D03")
write.csv(markers, file = paste0(run.name, "/", run.name,"_", compare,"_DE_genes.csv"))
  
top50 <- markers %>%
  slice_max(n = 50, order_by = avg_log2FC)
write.csv(top50, file = paste0(run.name, "/", run.name, "_", compare,"_top20_genes.csv"))
  
bottom50 <- markers %>%
  slice_min(n = 50, order_by = avg_log2FC)
write.csv(bottom50, file = paste0(run.name, "/", run.name,"_", compare,"_bottom20_genes.csv"))
}
```
