---
title: "LP"
output:
  pdf_document: default
  html_document: default
date: "2023-08-23, 20240210"
---

# try to repeat individuals from beggining by repeating 20240210-repeatIndividual.Rmd, colton's guide, and 20240212_integrateAll.Rmdusing new h5 generated on 20240202

Problem: couldn't RunAzimuth after integration due to layers

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(hdf5r)
library(dplyr)
library(reticulate)
library(DESeq2) # if want to perform DE using this option
library(reticulate)
library(BPCells)
library(tidyverse)
use_condaenv("/data/pengl7/conda/envs/r4seurat")
py_config()
# need to first create a conda env as r4seurat; conda active r4seurat; conda install leidenalg; conda install numpy pandas; start rstudio&
leidenalg <- import("leidenalg")
```

```{r}
sink("sesssionInfo_20240328.txt")
sessionInfo()
sink()
```

# file path for all samples

```{r}
h5_file_dir <- "../../raw_data/h5files"
merged_seurat_out_path <- "results/initial_merged_seurat_object.rds"
standard_integrated_seurat_out_path <- "results/final_integrated_seurat_object_using_full_dataset_integration.rds"
qc_data_output_path <- "results/initial_qc_metrics.tsv"
```

```{r}
# Get a list of all available h5 files
files_to_read_in <- list.files(
  path = h5_file_dir,
  pattern = "*h5",
  full.names = TRUE
)
```

#Error in curl::curl_fetch_memory(url, handle = handle) : 
  Timeout was reached: [www.ensembl.org:443] Operation timed out after 60001 milliseconds with 0 bytes received
```{r}
# Loop through h5 files
data_list <- c()
metadata <- c()

for (i in 1:length(files_to_read_in)) {
  
  path <- files_to_read_in[i]
  sample_name <- gsub(".h5$", "", basename(path))
  
  print(sample_name)
  
  # Write BP matrices
  data <- open_matrix_10x_hdf5(path)
  colnames(data) <- paste(colnames(data), sample_name, sep = "_")
  
  write_matrix_dir(
    mat = data,
    dir = paste0(gsub(".h5$", "", path), "_BP")
  )
  
  # Load in BP matrices
  mat <- open_matrix_dir(dir = paste0(gsub(".h5$", "", path), "_BP"))
  mat <- Azimuth:::ConvertEnsembleToSymbol(mat = mat, species = "mouse")
  
  # Create metadata
  temp_metadata <- data.frame(
    BarcodeID = colnames(mat),
    Sample = gsub(".h5$", "", basename(path))
  ) %>%
    mutate(
      Strain = gsub("-D[0-9]+$", "", Sample),
      TimePoint = str_extract(Sample, "D[0-9]+$")
    )
  
  # Add counts and metadata to their lists or dataframes
  data_list[[i]] <- mat
  metadata <- bind_rows(
    metadata,
    temp_metadata
  )
  
}
# Name layers
names(data_list) <- gsub(".h5", "", basename(files_to_read_in))

# Move add rownames to metadata
metadata <- metadata %>%
  column_to_rownames(var = "BarcodeID")

# Create merged seurat object
merged_seurat_object <- CreateSeuratObject(counts = data_list,
                                           meta.data = metadata)
#Save output ###

write_tsv(
  merged_seurat_object@meta.data,
  qc_data_output_path
)

# Save RDS object
saveRDS(
  object = merged_seurat_object,
  file = merged_seurat_out_path
)

```

### Pre-process raw merged Seurat Object ###
```{r}
# Load seurat object
merged_seurat_object <- readRDS(merged_seurat_out_path)

# Filter seurat object
merged_seurat_object <- subset(merged_seurat_object,
                               subset = nFeature_RNA > 300 &
                                 nFeature_RNA < 3000)

# Normalize data, find variable features, scale data, run PCA
merged_seurat_object <- NormalizeData(merged_seurat_object)
merged_seurat_object <- FindVariableFeatures(merged_seurat_object,
                                             verbose = FALSE)
merged_seurat_object <- ScaleData(merged_seurat_object)
merged_seurat_object <- RunPCA(merged_seurat_object)
```
# perform sctransform and annotation
```{r}
# Increase default globals size to ensure we don't exceed the default
sctransform_integrated_seurat_out_path <- "results/final_integrated_seurat_object_using_full_dataset_integration_and_sctransform.rds"

# Run SCTransform and PCA
merged_seurat_object <- SCTransform(merged_seurat_object, verbose = FALSE)
merged_seurat_object <- RunPCA(merged_seurat_object)

### Perform integration ###
# Increase default globals size to ensure we don't exceed the default
options(future.globals.maxSize = 24000 * 1024^2)
integrated_seurat_object <- IntegrateLayers(
  object = merged_seurat_object,
  method = RPCAIntegration,
  normalization.method = "SCT",
  new.reduction = "integrated.rpca",
  verbose = F
)

# Find clusters using leiden clustering
integrated_seurat_object <- FindNeighbors(
  object = integrated_seurat_object,
  reduction = "integrated.rpca",
  dims = 1:30
)

for(i in 1:length(seq(0.1, 0.5, 0.1))){
  
  integrated_seurat_object <- FindClusters(
    object = integrated_seurat_object,
    resolution = seq(0.1, 0.5, 0.1)[i], # Should explore and adjust resolution
    algorithm = 4,
    method = "igraph"
  )
  
}

integrated_seurat_object <- RunUMAP(
  object = integrated_seurat_object,
  reduction = "integrated.rpca",
  reduction.name = "umap.rpca",
  dims = 1:30, # Elbow plot confirmed this is an appropriate resolution
  return.model = T,
  verbose = F
)

```
Error in indices.ba[neighbors.ab, 1:k.anchor, drop = FALSE] : 
  subscript out of bounds
  
The "subscript out of bounds" error you're encountering with RunAzimuth might be due to having multiple layers in your Seurat object. A workaround suggested by a user who faced a similar issue is to use JoinLayers() or to subset each layer out as individual Seurat objects before running RunAzimuth. This approach resolved the problem for them.

```{r}
reference <- "../../hamsterLungRef" #"lungref"
integrated_seurat_object <- JoinLayers(integrated_seurat_object)
integrated_seurat_object <- RunAzimuth(integrated_seurat_object, reference = reference)

```
# took 1 hour to finish save and another 1 hour to release mem
```{r}
saveRDS(
  object = integrated_seurat_object,
  file = sctransform_integrated_seurat_out_path
)
```

#JoinLayers() is not working for SCT assay after integration
```{r}
sctransform_integrated_seurat_out_path <- "results/final_integrated_seurat_object_using_full_dataset_integration_and_sctransform.rds"
sctransform_integrated_seurat_object <- readRDS(sctransform_integrated_seurat_out_path)
```
