---
title: "20250402-EDA"
output: html_document
---

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(hdf5r)
library(dplyr)
library(reticulate)
library(DESeq2) # if want to perform DE using this option
library(reticulate)
library(BPCells)
library(tidyverse)
library(here)
use_condaenv("/data/pengl7/conda/envs/r4seurat")
py_config()
# need to first create a conda env as r4seurat; conda active r4seurat; conda install leidenalg; conda install numpy pandas; start rstudio&
leidenalg <- import("leidenalg")
```

```{r}
merged <- readRDS(here("results", "merged_filtered_percentile_filtered.rds"))
```


```{r}
merged
```

# Each sample has about 7K cells after initial data filtering

```{r}
head(merged@meta.data)
table(merged$orig.ident)  # or check other metadata columns like merged$condition
```
# add meta inforjmation: condition, replicate, batch 

```{r}
merged$condition <- ifelse(grepl("WT", merged$orig.ident), "WT", "KO")
merged$replicate <- ifelse(grepl("1", merged$orig.ident), "1", "2")
head(merged$condition)
```

# Basic QC plots
1. Before batch correction, check for potential variation across samples.
```{r}
VlnPlot(merged, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = "orig.ident", pt.size = 0.1)

```

# FeatureScatter
nCount_RNA VS nFeature_RNA
A strong correlation here is typical. Outliers (too many counts but few features) might be doublets or dead cells.


```{r}
FeatureScatter(merged, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

```{r}

```


# Normalize and Identify Variable Genes

```{r}
merged <- NormalizeData(merged)
merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 2000)
```

#. PCA, UMAP & Clustering
This is still pre-batch correction, just to explore structure:

```{r}
merged <- ScaleData(merged, vars.to.regress = "percent.mt")  # Optional regression
merged <- RunPCA(merged)
```


# check the difference by choosing pcs 30 and 50
```{r}
ElbowPlot(merged, ndims = 50)

```


```{r}
merged <- RunUMAP(merged, dims = 1:20)
DimPlot(merged, label = TRUE)

merged <- RunUMAP(merged, dims = 1:30)
DimPlot(merged, label = TRUE)

```
# chose PCs=30 to perform clustering

```{r}

merged <- RunUMAP(merged, dims = 1:30)
merged <- FindNeighbors(merged, dims = 1:30)
merged <- FindClusters(merged, resolution = 0.4)
```

Visualize clusters:
```{r}
DimPlot(merged, group.by = "seurat_clusters", label = TRUE)

```



```{r}
merged <- FindClusters(merged, resolution = 0.3)
DimPlot(merged, group.by = "seurat_clusters", label = TRUE)
```


```{r}
merged <- FindClusters(merged, resolution = 0.2)
DimPlot(merged, group.by = "seurat_clusters", label = TRUE)
```


```{r}
#install.packages("clustree")
library(clustree)
clustree(merged, prefix = "RNA_snn_res.")

```


# Assess cluster quality of r0.2

Need to join layers before performing FindAllMarkers since I merged 4 samples together which keep their indiviaul data and count slots

```{r}
merged <- JoinLayers(
  object = merged,
  assay = "RNA",
  slot = "data",                     # or "counts" if you're starting from raw counts
  new.layer = "joined_normalized"    # name for the new combined layer
)

```

Before JoinLayer
4 layers present: counts.1KO, counts.1WT, counts.2KO, counts.2WT
After JoinLayer: 3 layers present: data, counts, scale.data

```{r}
merged
```

Use marker genes to check whether these clusters represent distinct biological identities:

```{r}
Idents(merged) <- "seurat_clusters"
markers <- FindAllMarkers(
  merged,
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.1
)
```


```{r}
top10 <- markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
DoHeatmap(merged, features = top10$gene) + NoLegend()
```
visualize specific genes

```{r}
FeaturePlot(merged, features = c("Upk3a", "Acta2", "Col1a1", "Cd3e", "Cd79a"))
DotPlot(merged, features = c("Upk3a", "Acta2", "Col1a1", "Cd3e", "Cd79a")) + RotatedAxis()

```


```{r}
DotPlot(merged, features = list(
  Urothelial = c("Upk3a", "Krt20", "Krt8"),
  SmoothMuscle = c("Acta2", "Tagln", "Myh11"),
  Fibroblast = c("Col1a1", "Dcn", "Pdgfra"),
  Tcell = c("Cd3e", "Cd3d", "Trac"),
  Bcell = c("Cd79a", "Ms4a1"),
  Macrophage = c("Lyz2", "Adgre1")
)) + RotatedAxis()

```


```{r}
VlnPlot(merged, features = c("Acta2", "Col1a1", "Cd3e"), group.by = "seurat_clusters", pt.size = 0.1)

```

# Step-by-Step: Check for Batch Effects in Seurat
Plot UMAP by batch/sample/condition
```{r}
# Color by sample identity
DimPlot(merged, group.by = "orig.ident", label = TRUE, repel = TRUE)


```


```{r}

# Optional: split UMAP by sample or condition
DimPlot(merged, split.by = "orig.ident", group.by = "seurat_clusters")
```


```{r}
DimPlot(merged, split.by = "condition", group.by = "seurat_clusters")
```

# Visualize sample composition per cluster
This helps quantify how balanced each cluster is across samples:
```{r}
# Table of cluster × sample counts
tbl <- table(merged$seurat_clusters, merged$orig.ident)

# Reorder samples: WT1, WT2, KO1, KO2
sample_order <- c("1WT", "2WT", "1KO", "2KO")  # adjust based on your actual sample names
# Reorder the table columns
tbl <- tbl[, sample_order]
# Colors for clusters — use a consistent vector across plot & legend
cluster_colors <- 1:nrow(tbl)  # or you can define manually, like c("black", "red", ...)
# Proportions per sample
tbl_prop <- prop.table(tbl, margin = 2)

# Plot: That means 4 colors (recycled for each group of bars)
par(mar = c(5, 4, 4, 8), xpd = TRUE)
barplot(tbl_prop, beside = TRUE, col = 1:4,
        main = "Sample composition per cluster",
        ylab = "Proportion")

# Legend
legend("topright", inset = c(-0.2, 0),
       legend = rownames(tbl),
       fill = cluster_colors,
       title = "Cluster",
       cex = 0.8)
```
```{r}
# Normalize proportions per cluster
tbl_prop <- prop.table(tbl, margin = 2)

# Set layout so there's space on the right for the legend
par(mar = c(5, 4, 4, 8), xpd = TRUE)

# Barplot with room for the legend
barplot(tbl_prop,
        beside = TRUE,
        col = rainbow(nrow(tbl)),
        main = "Sample composition per cluster",
        ylab = "Proportion",
        las = 2)

# Add legend to the right
legend("topright", inset = c(-0.2, 0), legend = rownames(tbl),
       fill = rainbow(nrow(tbl)), title = "Cluster", cex = 0.8)
```


```{r}
as.data.frame(tbl) %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(x = "Cluster", y = "Proportion", fill = "Sample") +
  theme_minimal()

```

# save the rds which has been clustred and joinedLayer
```{r}
saveRDS(merged, file = here("results", "layerJoined_clusted.rds"))
```


```{r}
```


```{r}
```


```{r}
```

