---
title: "Untitled"
output: html_document
date: "2025-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# This is executed in GCP VM rstudio server

```{r}
# install.packages("remotes")
# remotes::install_github("carmonalab/UCell")
```

```{r}
library(Seurat)
library(here)
library(UCell)
library(dplyr)
library(ggplot2)
library(fs) # for dir_create()
library(patchwork) #for wrap figs

```

```{r}
set.seed(123)
```




# load the object

```{r}
merged <- readRDS(here("results", "final.rds"))
```

# create new output folder

```{r}
pathways_out <- "explore_pathways_output"
genes_out <- "explore_genes_output"
dir.create(here("results", pathways_out))
dir.create(here("results", genes_out))
```


# explore the marker expression of the pathways of interest using FeaturePlot and VlnPlot

Focus on smooth muscle cells, compare KO to WT

```{r}
pathways <- list(
  "Contractile_Protein" = c("Acta2", "Myocd", "Tagln", "Cald1", "Cnn1", "Des", "Vcl", "Myh11", "Smtn"),
  "RhoA_ROCK_signaling" = c("Rhoa", "Rock1", "Rock2"),
  "Calcium_signaling" = c("Mlck", "Camk2a", "Ppp3ca"),
  "GPCR_signaling" = c("Chrm2", "Chrm3", "Adrb1", "Adrb2", "Adrb3", "P2rx3", "P2ry1", "Gnaq", "Gnas"),
  "Prostaglandin_receptors" = c("Ptger1", "Ptger2", "Ptger3", "Ptger4", "Ptgfr"),
  "TGF_B_signaling" = c("Tgfb1", "Tgfb2", "Tgfbr1", "Tgfbr2"),
  "TRP_channels" = c("Trpv1", "Trpv4", "Trpc6", "Trpm4"),
  "EGFR_signaling" = c("Egfr", "Erbb2", "Erbb3", "Grb2", "Mapk1"),
  "Notch_Wnt_signaling" = c("Notch1", "Notch2", "Wnt1", "Wnt3a", "Ctnnb1", "Dll1"),
  "ECM_receptor_interaction" = c("Col1a1", "Col4a1", "Fn1", "Lama1", "Itga5", "Itgb1"),
  "TGF_B_SMAD_signaling" = c("Smad2", "Smad3", "Smad4", "Smad7"),
  "PI3K_AKT_signaling" = c("Pik3ca", "Pik3cb", "Akt1", "Akt2", "Mtor", "Pten"),
  "Wnt_signaling" = c("Wnt1", "Wnt3a", "Fzd7", "Lrp6", "Tcf7l2"),
  "Integrin_signaling" = c("Itgb1", "Itga5", "Tln1", "Vcl"),
  "AP-1_signaling" = c("Jun", "Junb", "Jund", "Fos", "Fosb", "Fosl1", "Fosl2" )
)


```


# explore the expression of the genes of interest individually
# first using Feature plot


```{r}
for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- FeaturePlot(merged, features = valid_genes, combine = FALSE)
    plots <- lapply(plots, function(p) p + theme(title = element_text(size = 10)))
    combined_plot <- wrap_plots(plots) + plot_annotation(title = pathway)
    
    tryCatch({
    ggsave(filename = here("results", genes_out, paste0(pathway, ".png")), plot = combined_plot, width = 12, height = 8, dpi = 300)
    })
    
  } else {
    message(paste("No valid genes found for:", pathway))
  }
}
  

```


```{r}
for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- FeaturePlot(merged, features = valid_genes, split.by = "condition")
    plots <- lapply(plots, function(p) p + theme(title = element_text(size = 10)))
    combined_plot <- wrap_plots(plots) + plot_annotation(title = pathway)
    
    tryCatch({
    ggsave(filename = here("results", genes_out, paste0(pathway, "_condition.png")), plot = combined_plot, width = 12, height = 8, dpi = 300)
    })
    
  } else {
    message(paste("No valid genes found for:", pathway))
  }
}
  
```


Use violin plot and break into either condition or cell types

```{r}
split_column <- "condition"

for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- VlnPlot(
      merged,
      features = valid_genes,
      split.by = split_column,
      pt.size = 0,
      combine = FALSE
    )
    
    plots <- lapply(plots, function(p) {
      p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    })
    
    combined_plot <- wrap_plots(plots) + plot_annotation(title = paste0(pathway, " (split by ", split_column, ")"))
    
      ggsave(
        filename = here("results", genes_out, paste0(pathway, "_violin_split.png")),
        plot = combined_plot, 
        width = 14, 
        height = length(valid_genes)*2,  # adjust height per number of plots
        dpi = 300
      )
      message(paste("Saved:", pathway, "split violin plot"))

  } else {
    message(paste("No valid genes found for:", pathway))
  }
}
```



```{r}
for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- VlnPlot(
      merged,
      features = valid_genes,
      group.by = "condition",
      pt.size = 0,
      combine = FALSE
    )
    
    plots <- lapply(plots, function(p) {
      p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    })
    
    combined_plot <- wrap_plots(plots) + plot_annotation(title = paste0(pathway, " (split by ", split_column, ")"))
    
      ggsave(
        filename = here("results", genes_out, paste0(pathway, "_violin_condition.png")),
        plot = combined_plot, 
        width = 14, 
        height = length(valid_genes)*2,  # adjust height per number of plots
        dpi = 300
      )
      message(paste("Saved:", pathway, "split violin plot"))

  } else {
    message(paste("No valid genes found for:", pathway))
  }
}

```



# Score pathways Using UCell

```{r}
merged <- AddModuleScore_UCell(merged, features = pathways)
```


```{r}
colnames(merged@meta.data)[grepl("_UCell$", colnames(merged@meta.data))]
```

```{r}
signature.names <- paste0(names(pathways), "_UCell")
signature.names
```

# No apprent difference on the signatures of interet on the pool of cells

```{r}
p <- VlnPlot(
  merged,
  features = signature.names,
  group.by = "condition",
  pt.size = 0
)

# Reduce title font size and x-axis text size
p <- p + theme(
  strip.text = element_text(size = 4),          # Smaller facet titles
  axis.text.x = element_text(size = 8, angle = 0, hjust = 0.5),
  axis.text.y = element_text(size = 8)
)

# Auto-adjust plot height and width
#n_features <- length(signature.names)
#plot_width <- 4 + 2 * (n_features >= 3)          # Scales width modestly with number of features
#plot_height <- max(4, n_features * 1.5)          # Height scales per number of violins

# Save the plot
ggsave(
  filename = here("results", pathways_out, "pathways_vlnplot_condition.png"),
  plot = p,
  width = 8,
  height = 12,
  dpi = 300
)
```


```{r}
p
```


```{r}
p <- VlnPlot(merged, features = signature.names, group.by = "sample", pt.size = 0)
ggsave(here("results", pathways_out,"pathways_vlnplot_sample.png"), 
       plot = p, 
       width = 8, 
       height = 12, 
       dpi = 300)
p
```


# subset cells of interest and apply the above
```{r}
# if has performed AddModuleScore function on the obj once, need to remove them before perform on the sub-obj
merged <- readRDS(here("results", "final.rds"))
```


```{r}
# Loop over each cell type

celltypes <- unique(merged$refined_cell_type)

for (ct in celltypes) {
  message("Processing cell type: ", ct)

  # Subset Seurat object by cell type
  subset_obj <- subset(merged, subset = refined_cell_type == ct)

  # Skip if too few cells
  if (ncol(subset_obj) < 50) {
    message("  Skipped: too few cells.")
    next
  }

  # Score pathway activity
  subset_obj <- AddModuleScore_UCell(subset_obj, features = pathways)
  signature.names <- paste0(names(pathways), "_UCell")
  # Create violin plot
  p <- VlnPlot(subset_obj, features = signature.names, group.by = "condition", pt.size = 0, ncol = 3)

  # Create output folder
  output_dir <- here("results", pathways_out, ct)
  dir_create(output_dir)

  # Save plot
  plot_file <- file.path(output_dir, "pathways_vlnplot_condition.png")
  ggsave(plot_file, plot = p, 
         width = 12, 
         height = 12, 
         dpi = 300)
}
```
# Plot the expression before smoothing UCell scores

```{r}
for (ct in celltypes) {
  message("Processing cell type: ", ct)

  # Subset Seurat object by cell type
  subset_obj <- subset(merged, subset = refined_cell_type == ct)

  # Skip if too few cells
  if (ncol(subset_obj) < 50) {
    message("  Skipped: too few cells.")
    next
  }

  # Score pathway activity
  subset_obj <- AddModuleScore_UCell(subset_obj, features = pathways)
  signature.names <- paste0(names(pathways), "_UCell")
  
  # Create output folder
  output_dir <- here("results", pathways_out, ct)
  dir_create(output_dir)
  
  # Determine number of panels (each feature × split)
  n_features <- length(signature.names)
  n_panels <- n_features * 2  # 2 panels per feature if split by condition
  ncol_plot <- 3
  nrow_plot <- ceiling(n_panels / ncol_plot)
  
  # Adjust width and height (e.g., 4x4 inches per panel)
  plot_width <- ncol_plot * 4
  plot_height <- nrow_plot * 4
  
  # Generate FeaturePlot
  p <- FeaturePlot(
    subset_obj,
    features = signature.names,
    split.by = "condition",
    pt.size = 0,
    ncol = ncol_plot
  )
  
  # Save the plot
  ggsave(
    filename = file.path(output_dir, "pathways_featurePlot_condition.png"),
    plot = p,
    width = plot_width,
    height = plot_height,
    dpi = 300
  )
}

```
key marker: Acta2 in the contractile signaling and Junb in the AP-1 signaling

```{r}
features = c("Acta2", "Junb")
VlnPlot(merged, features = "Junb", split.by = "condition")
VlnPlot(merged, features = "Acta2", split.by = "condition")
```


# AP-1 signal 
upregulated in endothelial-1, -2, fibroblast-1
downregulated in urotheilium, smooth muscle, and transitional cells

Contractile signal is down in endothial-1 and -2, transitional cells, and all urothelium cells


apply wilcoxon test 

✅ Step-by-Step: Statistical Testing of UCell Scores


```{r}

```

