---
title: "Untitled"
output: html_document
date: "2025-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# # Bioconductor core
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install(c("SingleR", "celldex", "SummarizedExperiment"))
# 
# # Optional: for converting Seurat -> SingleCellExperiment
# install.packages("SingleCellExperiment")

```


```{r}
library(SingleR)
library(celldex)
library(Seurat)
library(SingleCellExperiment)
library(here)
```


```{r}
merged <- readRDS(here("results", "layerJoined_clusted.rds"))
DefaultAssay(merged)
```


```{r}
DefaultLayer(merged[["RNA"]]) <- "data" 
```


```{r}
Idents(merged) <- "seurat_clusters"
```

Re-cluster Using the Leiden Algorithm
 Leiden clustering in Seurat depends on the Python leidenalg package, which in turn requires Python and pandas to be available via reticulate (R’s bridge to Python).

```{r}
library(reticulate)

# See which Python environment R is using
py_config()

```

Install pandas into your existing virtualenv
```{r}
# reticulate::use_virtualenv("/spin1/home/linux/pengl7/.virtualenvs/r-reticulate", required = TRUE)
# reticulate::py_install("pandas")

```

```{r}
# (Re-)run FindNeighbors if not already done
merged <- FindNeighbors(merged, dims = 1:30)


```

Running leiden takes lots mem > 80G, max use 100G for 2.1G rds object
```{r}

# Run Leiden clustering
merged <- FindClusters(merged, algorithm = 4, resolution = 0.2)  # you can adjust resolution
```


There isn't big difference on the clustering beteeen leiden and default
```{r}

# Check new clusters
table(merged$seurat_clusters)
DimPlot(merged, group.by = "seurat_clusters", label = TRUE, repel = TRUE)
```


```{r}
saveRDS(merged, here("results", "clusterByLeiden0.2.rds"))
```

# try annotation using singleR

Convert Your Seurat Object to SingleCellExperiment
Make sure the correct assay and layer are set:
```{r}
DefaultAssay(merged) <- "RNA"
DefaultLayer(merged[["RNA"]]) <- "data"

# Convert
sce <- as.SingleCellExperiment(merged)
```

There’s no reference called “bladder” specifically, but ImmGen and MouseRNAseq contain many of the relevant immune, stromal, and epithelial types you’d expect in bladder tissue.

Try this first:
```{r}
ref <- celldex::ImmGenData(verbose = TRUE)  # good for immune + stromal cells

```

```{r}
saveRDS(ref, "ImmGenData.rds")
```


Run SingleR Annotation
```{r}
# This will return a data frame with predicted labels and confidence scores for each cell.
pred <- SingleR(test = sce, ref = ref, labels = ref$label.main)
# If you want more granular labels:
# pred <- SingleR(test = sce, ref = ref, labels = ref$label.fine)

```

Add Labels Back to Seurat Object
```{r}
merged$SingleR <- pred$labels
Idents(merged) <- "SingleR"
DimPlot(merged, group.by = "SingleR", label = TRUE, repel = TRUE)

```

```{r}
str(pred)

```


```{r}
delta_values <- as.numeric(pred$delta.next)
hist(delta_values, breaks = 50,
     main = "SingleR Prediction Confidence",
     xlab = "Delta (top vs second-best score)")

```


# check the cell doublets and score cell cycle

The expected doublet rate for 10x Genomics data is strongly dependent on the number of cells loaded during GEM (Gel Bead-in-Emulsion) generation, and 10x has published fairly detailed guidance.

10K cells per sample Expected Doublet Rate is ~8.0%
Or use a slightly conservative estimate like 5–6% globally

```{r}
#devtools::install_github('chris-mcginnis-ucsf/DoubletFinder')
library(DoubletFinder)
```



```{r}
# Estimate expected doublets: ~1% for 10x, or use nExp = 0.05 * ncol(merged)
nExp_poi <- round(0.05 * ncol(merged))  # 5% doublets

# Run DoubletFinder
sweep.res <- paramSweep(merged, PCs = 1:30, sct = FALSE)
sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
bcmvn <- find.pK(sweep.stats)
```
```{r}
library(ggplot2)

# Convert pK column to numeric (if needed)
bcmvn$pK <- as.numeric(as.character(bcmvn$pK))

# Find pK with highest BCmetric
best_index <- which.max(bcmvn$BCmetric)
best_pK <- bcmvn$pK[best_index]
best_metric <- bcmvn$BCmetric[best_index]

cat("Best pK:", best_pK, "with BCmetric:", best_metric, "\n")

# Plot with labeled axes and a line at best pK
ggplot(bcmvn, aes(x = pK, y = BCmetric)) +
  geom_point(color = "dodgerblue") +
  geom_line(color = "dodgerblue") +
  geom_vline(xintercept = best_pK, color = "red", linetype = "dashed") +
  annotate("text", x = best_pK, y = best_metric + 0.01, label = paste0("Best pK = ", best_pK),
           hjust = 0, color = "red", size = 3) +
  labs(
    title = "BCmetric vs pK",
    x = "pK (neighborhood size)",
    y = "BCmetric (classification performance)"
  ) +
  theme_minimal()


```

pN: This is almost always left at 0.25, the recommended default: 0.25
pK: choose according to the above BCmetric plot

```{r}

best_pK <- as.numeric(bcmvn$pK[which.max(bcmvn$BCmetric)])

# Convert seurat_clusters column to character
#merged$seurat_clusters <- as.character(merged$seurat_clusters)
#Idents(merged) <- merged$seurat_clusters

merged <- doubletFinder(
  merged,
  PCs = 1:30,
  pN = 0.25,
  pK = best_pK,        # ← use your top pK here
  nExp = round(0.08 * ncol(merged)),
  sct = FALSE
)

```


```{r}
# Inspect results
table(merged@meta.data[[grep("DF.classification", names(merged@meta.data), value = TRUE)]])
DimPlot(merged, group.by = grep("DF.classification", names(merged@meta.data), value = TRUE))
```


Compare to the estimate doublet rate of 4%

```{r}
merged <- doubletFinder(
  merged,
  PCs = 1:30,
  pN = 0.25,
  pK = best_pK,        # ← use your top pK here
  nExp = round(0.04 * ncol(merged)),
  sct = FALSE
)
```


```{r}
# Inspect results
table(merged@meta.data[[grep("DF.classification", names(merged@meta.data), value = TRUE)]])
DimPlot(merged, group.by = grep("DF.classification", names(merged@meta.data), value = TRUE))
```


```{r}
```

