---
title: "LP"
output:
  pdf_document: default
  html_document: default
date: "2025-03-28"
---

# load required packages following 20240328_followColton.Rmd
R --version is 4.3.2

```{r}
.libPaths()
```

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(Azimuth)
library(sctransform)
library(ggplot2)
library(hdf5r)
library(dplyr)
library(reticulate)
library(DESeq2) # if want to perform DE using this option
library(reticulate)
library(BPCells)
library(tidyverse)
library(here)
use_condaenv("/data/pengl7/conda/envs/r4seurat")
py_config()
# need to first create a conda env as r4seurat; conda active r4seurat; conda install leidenalg; conda install numpy pandas; start rstudio&
leidenalg <- import("leidenalg")
```
```{r}
find.package("Seurat")
# "/usr/local/apps/R/4.3/site-library_4.3.2/Seurat"
```

```{r}
DATE="20250328"
sink(paste0("sesssionInfo_", DATE, ".txt"))
sessionInfo()
sink()
```

# file path for all samples
```{r}
library(here)
```

```{r}
# Directory containing multiple sample folders
h5_file_dir <- here("data")
rds_dir <- here("results", "rds_unfiltered")
dir.create(rds_dir, showWarnings = FALSE)

# Path to save Seurat object and outputs
merged_out_path <- here("results", "merged.rds") 
integrated_out_path <- here( "results", "integrated.rds")
qc_data_output_path <- here("results", "initial_qc_metrics.tsv")
```

```{r}
# Discover sample folders inside h5_file_dir
sample_dirs <- list.dirs(h5_file_dir, recursive = FALSE, full.names = TRUE)
print(sample_dirs)
sample_names <- basename(sample_dirs)
print(sample_names)
```

```{r}
# Initialize list to store individual Seurat objects
seurat_list <- list()
qc_metrics_list <- list()

# Loop through samples
for (i in seq_along(sample_dirs)) {
  sample <- sample_names[i]
  h5_file <- file.path(sample_dirs[i], "filtered_feature_bc_matrix.h5")

  message("🔄 Reading: ", h5_file)
  counts <- Read10X_h5(h5_file)
  seurat_obj <- CreateSeuratObject(counts = counts, project = sample, min.cells = 3, min.features = 200)

  # Annotate with metadata
  seurat_obj$sample <- sample
  seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt-")  # adjust to "^MT-" for human

  # Store QC info
  qc_df <- seurat_obj@meta.data %>% mutate(cell = colnames(seurat_obj), sample = sample)
  qc_metrics_list[[sample]] <- qc_df

  # Save Seurat object as individual .rds
  rds_path <- file.path(rds_dir, paste0(sample, "_raw.rds"))
  saveRDS(seurat_obj, file = rds_path)

  # Add to list for later merging
  seurat_list[[sample]] <- seurat_obj
}

# Combine all QC metrics
all_qc <- bind_rows(qc_metrics_list)
write.table(all_qc, file = qc_output_path, sep = "\t", quote = FALSE, row.names = FALSE)

# Optional: Save full unfiltered list
# seurat_list <- readRDS("results/seurat_list_unfiltered.rds") if want to read the list
saveRDS(seurat_list, file = here("results", "seurat_list_unfiltered.rds"))

message("✅ Done! Raw Seurat objects and QC metrics saved.")
```


```{r}
qc_data_output_path <- here("results", "initial_qc_metrics.tsv")
data <- read_tsv(qc_data_output_path, show_col_types = FALSE)

lower_cutoff <- quantile(data$nFeature_RNA, probs = 0.05)
upper_cutoff <- quantile(data$nFeature_RNA, probs = 0.95)

p1 <- data %>%
  ggplot(aes(x = sample, y = nFeature_RNA)) +
  geom_violin(fill = "red", alpha = 0.5) +
  geom_jitter(alpha = 0.2, size = 0.2) +
  geom_hline(yintercept = lower_cutoff, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = upper_cutoff, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  ) +
  labs(x = NULL, y = "# of Detected Genes")

# Save the plot to PDF
ggsave(
  filename = here::here("results", "violin_nFeature.pdf"),
  plot = p1,
  width = 8,
  height = 5
)

message("✅ Violin plot for percent.mt saved as 'results/violin_nFeature.pdf'")

p1
```

# contiue on 20250329
read in the list of rds

```{r}
# Load unfiltered Seurat list
seurat_list <- readRDS(here("results", "seurat_list_unfiltered.rds"))
```

```{r}
seurat_list
```

```{r}
# Create the violin plot for mitochondrial percentage
#upper_cutoff <- quantile(data$percent.mt, probs = 0.90)
upper_cutoff <- 10
p3 <- data %>%
  ggplot(aes(x = sample, y = percent.mt)) +
  geom_violin(fill = "skyblue", alpha = 0.5) +
  geom_jitter(alpha = 0.2, size = 0.2) +
  geom_hline(yintercept = upper_cutoff, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  ) +
  labs(
    x = NULL,
    y = "Mitochondrial Percentage",
    title = "Mitochondrial Content per Sample"
  )

# Save the plot to PDF
ggsave(
  filename = here::here("results", "violin_percent_mt.pdf"),
  plot = p3,
  width = 8,
  height = 5
)

message("✅ Violin plot for percent.mt saved as 'results/violin_percent_mt.pdf'")

p3
```


```{r}
# Create the violin plot for total UMI count per cell
# Calculate dynamic thresholds
lower_cutoff <- quantile(data$nCount_RNA, probs = 0.05)
upper_cutoff <- quantile(data$nCount_RNA, probs = 0.95)

p2 <- data %>%
  ggplot(aes(x = sample, y = nCount_RNA)) +
  geom_violin(fill = "forestgreen", alpha = 0.5) +
  geom_jitter(alpha = 0.2, size = 0.2) +
  geom_hline(yintercept = lower_cutoff, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = upper_cutoff, linetype = "dashed", color = "red") + # optional guideline
  coord_flip() +
  theme_bw() +
  theme(
    axis.title = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  ) +
  labs(
    x = NULL,
    y = "Total UMI Counts (nCount_RNA)",
    title = "UMI Counts per Cell per Sample"
  )

# Save the plot to PDF
ggsave(
  filename = here::here("results", "violin_nCount_RNA.pdf"),
  plot = p2,
  width = 8,
  height = 5
)

message("✅ Violin plot for nCount_RNA saved as 'results/violin_nCount_RNA.pdf'")

p2
```

Common QC thresholds:as a default starting point, especially for human PBMC or similar datasets
subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5

For mouse tissues, including solid organs (like bladder), thresholds are often adjusted higher:
subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 10

Most studies primarily filter on:
nFeature_RNA
  Low values: empty droplets or poor quality
  High values: likely doublets or multiplets

percent.mt
  High mito % suggests damaged or apoptotic cells
  
Percentage of total reads that map to cells rather than background (ambient) RNA.
59–61% is lower than the ideal
Range (%)	Interpretation
>80%	Excellent capture, low ambient RNA
70–80%	Acceptable
60–70%	Borderline — check for ambient/damaged cells
<60%	High ambient RNA or poor capture
So you're right — it could indicate:

A large population of dying/low-RNA-content cells
Ambient RNA (especially from lysed cells)
Low complexity libraries

```{r}
# Define the percentiles you want
percentiles <- c(0.01, 0.05, 0.10, 0.90, 0.95, 0.99)

# Compute percentiles for each metric
result <- tibble(
  Percentile = paste0(percentiles * 100, "th"),
  nFeature_RNA = quantile(data$nFeature_RNA, probs = percentiles, na.rm = TRUE),
  nCount_RNA = quantile(data$nCount_RNA, probs = percentiles, na.rm = TRUE),
  percent.mt = quantile(data$percent.mt, probs = percentiles, na.rm = TRUE)
)

# View the results
print(result)

# Optionally write to file
write_tsv(result, here::here("results", "qc_percentiles.tsv"))
```


```{r}
# Extract all metadata to compute percentiles
all_meta <- bind_rows(
  lapply(seurat_list, function(obj) {
    obj@meta.data %>% mutate(cell = colnames(obj))
  })
)

# Calculate dynamic thresholds
lower_nFeature <- quantile(all_meta$nFeature_RNA, 0.05)
upper_nFeature <- quantile(all_meta$nFeature_RNA, 0.95)
max_percent_mt <- 12.5

message("🔎 Using thresholds: nFeature_RNA > ", round(lower_nFeature), 
        " & < ", round(upper_nFeature),
        " | percent.mt < ", max_percent_mt)

# Apply filtering to each sample
filtered_list <- lapply(seurat_list, function(obj) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^mt-")
  subset(obj, 
         subset = nFeature_RNA > lower_nFeature &
                  nFeature_RNA < upper_nFeature &
                  percent.mt < max_percent_mt)
})

# Merge all filtered Seurat objects
merged_filtered <- merge(
  filtered_list[[1]],
  y = filtered_list[-1],
  add.cell.ids = names(filtered_list),
  project = "HDAC11_Filtered"
)

# Save merged filtered object
saveRDS(merged_filtered, here("results", "merged_filtered_percentile_filtered.rds"))

message("✅ Filtering complete. Final merged object saved to: results/merged_filtered_percentile_filtered.rds")
```

# generate a plot function, them compare metrics before and after filtering
```{r}

# Extract metadata for violin plotting
meta_filtered <- merged_filtered@meta.data %>%
  mutate(cell = colnames(merged_filtered),
         sample = sample,            # Already assigned in previous steps
         status = "Post-filter")

# Function to plot a QC metric
plot_violin_metric <- function(df, metric, y_label, filename) {
  p <- df %>%
    ggplot(aes(x = sample, y = .data[[metric]], fill = status)) +
    geom_violin(alpha = 0.5, position = position_dodge(width = 0.8)) +
    geom_jitter(alpha = 0.2, size = 0.2, position = position_jitterdodge()) +
    coord_flip() +
    labs(x = NULL, y = y_label, title = paste("QC Metric:", y_label)) +
    theme_bw() +
    theme(axis.title = element_text(size = 14),
          axis.text = element_text(size = 12),
          legend.position = "top")

  ggsave(filename = here("results", filename), plot = p, width = 8, height = 5)
}

# Example: Plot only post-filter metrics
plot_violin_metric(meta_filtered, "nFeature_RNA", "# of Detected Genes", "violin_nFeature_post_filter.pdf")
plot_violin_metric(meta_filtered, "percent.mt", "Mitochondrial Percentage", "violin_percent_mt_post_filter.pdf")
```




```{r}
filtered <- subset(
  merged_seurat,
  subset =
    nFeature_RNA > 300 &
    nFeature_RNA < 6000 &
    nCount_RNA > 500 &
    nCount_RNA < 50000 &
    percent.mt < 12.5
)

```


```{r}
### How many cells per sample remaining after filtering ###

data %>%
  mutate(
    Status = ifelse(nFeature_RNA < 300 | nFeature_RNA > 6000, "Remove", "Keep")
  ) %>%
  group_by(sample, Status) %>%
  summarise(Count = n(),
            .groups = "drop") %>%
  pivot_wider(names_from = Status,
              values_from = Count)
```


