---
title: "Untitled"
output: html_document
date: "2025-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# This is executed in GCP VM rstudio server

```{r}
# install.packages("remotes")
# remotes::install_github("carmonalab/UCell")
```

```{r}
library(Seurat)
library(here)
library(UCell)
library(dplyr)
library(ggplot2)
library(fs) # for dir_create()
library(patchwork) #for wrap figs

```

```{r}
set.seed(123)
```




# load the object

```{r}
merged <- readRDS(here("results", "final.rds"))
```

# create new output folder

```{r}

# name the subfolder
feature_p <- "feature_plot"
violin_p <- "violin_plot"

# Generate the folders for individual gene exploration
#genes_out <- "explore_genes_output_20250915"
#dir.create(here("results", genes_out))

# for (subfolder in c(feature_p, violin_p)){
#   dir.create(here("results", genes_out, subfolder))
# }


# generate folders for pathway exploration
pathways_out <- "explore_pathways_output_20250915"
dir.create(here("results", pathways_out))

for (subfolder in c(feature_p, violin_p)){
  dir.create(here("results", pathways_out, subfolder))
  
}
```


# explore the marker expression of the pathways of interest using FeaturePlot and VlnPlot

Focus on smooth muscle cells, compare KO to WT

Lipoxygenases: (脂氧合酶LOX family) Alox5, Alox15, Alox12, Lta4h
Cyclooxygenases (环氧合酶 COX family) Ptgs1 Ptgs2
Cyp450 细胞色素 P450 超家族: 大型单加氧酶家族, it is a superfamily
Cyp1a1, Cyp1a2, Cyp2e1, Cyp2c29, Cyp2d22, Cyp3a11, Cyp4a10/14
Gravin-α: Akap12（蛋白激酶 A 锚定蛋白 12）: 信号支架蛋白，结合 PKA、PKC 和 Src 激酶

```{r}

pathways <- list(
  "Lox" = c("Alox5", "Alox15", "Alox12", "Lta4h"),
  "Cox" = c("Ptgs1", "Ptgs2"),
  "Cyp450" = c("Cyp1a1", "Cyp1a2", "Cyp2e1", "Cyp2c29", "Cyp2d22", "Cyp3a11", "Cyp4a10", "Cyp4a14"),
  "Garvin-alpha" = c("Akap12")
)

```


# explore the expression of the genes of interest using Feature plot

Only 3 Cyp genes were in th gene features

```{r}
for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  output_folder <- pathways_out
  output_subfolder <- feature_p
  
  if (length(valid_genes) > 0) {
    plots <- FeaturePlot(merged, features = valid_genes, combine = FALSE)
    plots <- lapply(plots, function(p) p + theme(title = element_text(size = 10)))
    combined_plot <- wrap_plots(plots) + plot_annotation(title = pathway)
    
    tryCatch({
    ggsave(filename = here("results", output_folder, output_subfolder, paste0(pathway, ".png")), plot = combined_plot, width = 12, height = 8, dpi = 300)
    })
    
  } else {
    message(paste("No valid genes found for:", pathway))
  }
}
  

```




```{r}
merged$refined_cell_type
```
# plot Akap12 seperately, since the outuput form below cause bad height
ggsave(
        filename = here(subfolder2, paste0(pathway, "_violin_celltype_by_", split_column, ".png")),
        plot = combined_plot, 
        width = 14, 
        height = length(valid_genes)*2,  # adjust height per number of plots when weird output
        dpi = 300
      )


```{r}
VlnPlot(merged, features = "Akap12",
        group.by = "refined_cell_type", split.by = "condition",
        assay = "RNA", slot = "data") + NoLegend()

```


Use violin plot and break into KO/WT condition per cell type

```{r}
split_column <- "condition"
output_folder <- pathways_out
output_subfolder <- violin_p
subfolder2 <- here("results", output_folder, output_subfolder, paste0("perCelltype_by", split_column))
dir.create(subfolder2)
    
for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- VlnPlot(
      merged,
      features = valid_genes,
      split.by = split_column, # here still grouped by cell types but then split further by condition
      pt.size = 0.02,
      combine = FALSE
    )
    
    plots <- lapply(plots, function(p) {
      p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    })
    
    combined_plot <- wrap_plots(plots) + 
      plot_annotation(
        title = paste0(pathway, " (split by ", split_column, ")"),
        theme = theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))  # <- left pad
)
  
      ggsave(
        filename = here(subfolder2, paste0(pathway, "_violin_celltype_by_", split_column, ".png")),
        plot = combined_plot, 
        width = 16, 
        height = length(valid_genes)*2,  # adjust height per number of plots when weird output
        dpi = 300
      )
      message(paste("Saved:", pathway, "split violin plot"))

  } else {
    message(paste("No valid genes found for:", pathway))
  }
}
```

# plot Akap12 seperately, since the outuput form below cause bad height
ggsave(
        filename = here(subfolder2, paste0(pathway, "_violin_celltype_by_", split_column, ".png")),
        plot = combined_plot, 
        width = 14, 
        height = length(valid_genes)*2,  # adjust height per number of plots when weird output
        dpi = 300
      )
      
```{r}
for (pathway in "Garvin-alpha") {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- VlnPlot(
      merged,
      features = valid_genes,
      split.by = split_column, # here still grouped by cell types but then split further by condition
      pt.size = 0.05,
      combine = FALSE
    )
    
    plots <- lapply(plots, function(p) {
      p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    })
    
    combined_plot <- wrap_plots(plots) + 
      plot_annotation(
        title = paste0(pathway, " (split by ", split_column, ")"),
        theme = theme(plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 50))  # <- left pad
)
  
      ggsave(
        filename = here(subfolder2, paste0(pathway, "_violin_celltype_by_", split_column, ".png")),
        plot = combined_plot, 
        width = 10, 
        height = 6,  # adjust height per number of plots when weird output
        dpi = 300
      )
      message(paste("Saved:", pathway, "split violin plot"))

  } else {
    message(paste("No valid genes found for:", pathway))
  }
}

```


############## Not execute below yet

# violin plot per condition for the whole cell pools
usually no big difference between conditions overall

```{r}
subfolder2 <- here("results", genes_out, violin_p, "overall_by_condition")
dir.create(subfolder2)

for (pathway in names(pathways)) {
  genes <- pathways[[pathway]]
  valid_genes <- genes[genes %in% rownames(merged)]
  
  if (length(valid_genes) > 0) {
    plots <- VlnPlot(
      merged,
      features = valid_genes,
      group.by = "condition", # change the grouping of cell type to condition
      pt.size = 0,
      combine = FALSE
    )
    
    plots <- lapply(plots, function(p) {
      p + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    })
    
    combined_plot <- wrap_plots(plots) + plot_annotation(title = paste0(pathway, " (split by condition) "))
    
      ggsave(
        filename = here(subfolder2, paste0(pathway, "_violin_overall_by_condition.png")),
        plot = combined_plot, 
        width = 14, 
        height = length(valid_genes)*2,  # adjust height per number of plots
        dpi = 300
      )
      message(paste("Saved:", pathway, "split violin plot"))

  } else {
    message(paste("No valid genes found for:", pathway))
  }
}

```


######################################
# Score pathways Using UCell and plot the expession scores

```{r}
merged <- AddModuleScore_UCell(merged, features = pathways)
```


```{r}
colnames(merged@meta.data)[grepl("_UCell$", colnames(merged@meta.data))]
```

```{r}
signature.names <- paste0(names(pathways), "_UCell")
signature.names
```

# No apprent difference on the signatures of interest on the pool of cells

```{r}
p <- VlnPlot(
  merged,
  features = signature.names,
  group.by = "condition",
  pt.size = 0
)

# Reduce title font size and x-axis text size
p <- p + theme(
  strip.text = element_text(size = 4),          # Smaller facet titles
  axis.text.x = element_text(size = 8, angle = 0, hjust = 0.5),
  axis.text.y = element_text(size = 8)
)

# Auto-adjust plot height and width
#n_features <- length(signature.names)
#plot_width <- 4 + 2 * (n_features >= 3)          # Scales width modestly with number of features
#plot_height <- max(4, n_features * 1.5)          # Height scales per number of violins

# Save the plot
ggsave(
  filename = here("results", pathways_out, "pathways_vlnplot_condition.png"),
  plot = p,
  width = 8,
  height = 12,
  dpi = 300
)
```


```{r}
p
```


# not executed this time
```{r}
p <- VlnPlot(merged, features = signature.names, group.by = "sample", pt.size = 0)
ggsave(here("results", pathways_out,"pathways_vlnplot_sample.png"), 
       plot = p, 
       width = 8, 
       height = 12, 
       dpi = 300)
p
```


####################
# subset cells of interest and apply the above UCscore
```{r}
# if has performed AddModuleScore function on the obj once, need to remove them before perform on the sub-obj
merged <- readRDS(here("results", "final.rds"))
```


# Plot the expression before smoothing UCell scores using violin plot

```{r}
# Loop over each cell type

celltypes <- unique(merged$refined_cell_type)

for (ct in celltypes) {
  message("Processing cell type: ", ct)

  # Subset Seurat object by cell type
  subset_obj <- subset(merged, subset = refined_cell_type == ct)

  # Skip if too few cells
  if (ncol(subset_obj) < 50) {
    message("  Skipped: too few cells.")
    next
  }

  # Score pathway activity
  subset_obj <- AddModuleScore_UCell(subset_obj, features = pathways)
  signature.names <- paste0(names(pathways), "_UCell")
  # Create violin plot
  p <- VlnPlot(subset_obj, features = signature.names, group.by = "condition", pt.size = 0, ncol = 3)

  # Create output folder
  output_dir <- here("results", pathways_out, ct)
  dir_create(output_dir)

  # Save plot
  plot_file <- file.path(output_dir, "pathways_vlnplot_condition.png")
  ggsave(plot_file, plot = p, 
         width = 12, 
         height = 12, 
         dpi = 300)
}
```

# Plot the expression using feature plot
didn't perform since feature plots are usally not as good as violin plot

```{r}
for (ct in celltypes) {
  message("Processing cell type: ", ct)

  # Subset Seurat object by cell type
  subset_obj <- subset(merged, subset = refined_cell_type == ct)

  # Skip if too few cells
  if (ncol(subset_obj) < 50) {
    message("  Skipped: too few cells.")
    next
  }

  # Score pathway activity
  subset_obj <- AddModuleScore_UCell(subset_obj, features = pathways)
  signature.names <- paste0(names(pathways), "_UCell")
  
  # Create output folder
  output_dir <- here("results", pathways_out, ct)
  dir_create(output_dir)
  
  # Determine number of panels (each feature × split)
  n_features <- length(signature.names)
  n_panels <- n_features * 2  # 2 panels per feature if split by condition
  ncol_plot <- 3
  nrow_plot <- ceiling(n_panels / ncol_plot)
  
  # Adjust width and height (e.g., 4x4 inches per panel)
  plot_width <- ncol_plot * 4
  plot_height <- nrow_plot * 4
  
  # Generate FeaturePlot
  p <- FeaturePlot(
    subset_obj,
    features = signature.names,
    split.by = "condition",
    pt.size = 0,
    ncol = ncol_plot
  )
  
  # Save the plot
  ggsave(
    filename = file.path(output_dir, "pathways_featurePlot_condition.png"),
    plot = p,
    width = plot_width,
    height = plot_height,
    dpi = 300
  )
}

```



apply wilcoxon test: not exeuted yet

✅ Step-by-Step: Statistical Testing of UCell Scores


# take a closer look at immune cells. 

# calculate the cell number

```{r}
table(merged$refined_cell_type)
```

```{r}
lymphoid <- subset(merged, subset = refined_cell_type == "Immune_Lymphoid")
```

```{r}
table(lymphoid$condition)
```


```{r}
myeloid <- subset(merged, subset = refined_cell_type == "Immune_Myeloid")
table(myeloid$condition)
```


# check the cell numbers among cell types
```{r}
for (ct in merged$refined_cell_type)
```


```{r}
unique(merged@meta.data$refined_cell_type)
```
```{r}
table(merged@meta.data$refined_cell_type, merged@meta.data$condition)
```
```{r}
cell_counts <- as.data.frame(table(merged@meta.data$refined_cell_type, merged@meta.data$condition))
cell_counts
```

