---
title: "Untitled"
output: html_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Seurat)
library(SingleR)
library(celldex)
library(here)
library(ggplot2)
library(dplyr)
```


```{r}
merged <- readRDS(here("results", "noDoublets_scoreCellCycle.rds"))
```

```{r}
names(merged@meta.data)
```

```{r}
DefaultAssay(merged) <- "RNA"
DefaultLayer(merged[["RNA"]]) <- "data"
Idents(merged) <- "seurat_clusters"
```


```{r}
DimPlot(merged, group.by = "seurat_clusters", label = TRUE, repel = TRUE)
```


```{r}
# Convert
sce <- as.SingleCellExperiment(merged)
#ref <- celldex::ImmGenData(verbose = TRUE)  # good for immune + stromal cells
#saveRDS(ref, "ImmGenData.rds")
#I already cache and save it.
ref <- readRDS(here("ref", "ImmGenData.rds"))
```

# first use the dataset of ImmGenData ref

```{r}
# This will return a data frame with predicted labels and confidence scores for each cell.
pred <- SingleR(test = sce, ref = ref, labels = ref$label.main)
# If you want more granular labels:
# pred <- SingleR(test = sce, ref = ref, labels = ref$label.fine)
```

Add Labels Back to Seurat Object
Use cell-level annotation to get a fine-grained view
if add clusters=seurat_obj$seurat_clusters, will perform cluster-level annotion which is faster, less noisy, but may miss heterogenity within a cluster and not useful if your clusters are mixed

```{r}
merged$SingleR_main <- pred$labels
Idents(merged) <- "SingleR_main"
DimPlot(merged, group.by = "SingleR_main", label = TRUE, repel = TRUE)
```


The delta score is the difference between the top and second-best predicted labels for each cell.

Higher delta = higher confidence.
Lower delta (< 0.2) means the classifier was not very confident — the top two labels were close in score.

If most deltas are < 0.2:

The predictions are low-confidence overall.
There might be mixed or ambiguous cell types, or
The reference dataset (ImmGenData) may not fully capture your cell types (especially if you have non-immune cells in your bladder dataset).

```{r}
delta_values <- as.numeric(pred$delta.next)
hist(delta_values, breaks = 50,
     main = "SingleR Prediction Confidence",
     xlab = "Delta (top vs second-best score)")

```

# test annother ref 
```{r}
# Load the MouseRNAseqData reference
ref_mouse <- celldex::MouseRNAseqData()
```


```{r}
saveRDS(ref_mouse, here("ref", "MouseRNAseqData"))
```


```{r}
# Run SingleR with this reference (cell-level annotation)
pred_mouse <- SingleR(test = sce,
                      ref = ref_mouse,
                      labels = ref_mouse$label.main)

merged$SingleR_mouse_main <- pred_mouse$labels
Idents(merged) <- "SingleR_mouse_main"
DimPlot(merged, group.by = "SingleR_mouse_main", label = TRUE, repel = TRUE)
```

# The results is worse

```{r}
delta_mouse_values <- as.numeric(pred_mouse$delta.next)
hist(delta_mouse_values, breaks = 50,
     main = "SingleR Prediction Confidence",
     xlab = "Delta (top vs second-best score)")
```
# try mannual annotation

```{r}
# One by one or a few at a time
seurat_obj <- merged
FeaturePlot(seurat_obj, features = c("Upk1a", "Krt5", "Krt20"))
FeaturePlot(seurat_obj, features = c("Pecam1", "Cdh5"))
VlnPlot(seurat_obj, features = c("Col1a1", "Acta2"))

```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

