---
title: "Untitled"
output: html_document
date: "2023-05-01"
---

## Sometimes have problem with installing packages, they are many ways to install packages
# here is an example for install seurat-data
# although it is finally installed successfuly. The accessing to lungref still failed

# not working
```{r}
install.packages(seurat-data) # not work
install.packages("https://api.github.com/repos/satijalab/seurat-data/tarball/seurat5") # not work
```

# not working; installing from git repo
```{r}
install_github("satijalab/seurat-data") # not work
remotes::install_github("satijalab/seurat-data", "seurat5") # not work
devtools::install_github('satijalab/seurat-data') # not work
```

# not working
```{r solution 1}
library(httr)
url <- "https://github.com/satijalab/seurat-data/archive/refs/heads/master.zip"
temp <- tempfile()
GET(url, write_disk(temp))
install.packages(temp, repos = NULL, type="source")
```

## If the above doesn't work, 
Go to the SeuratData GitHub repository at https://github.com/satijalab/seurat-data
Click on the green "Code" button and select "Download ZIP"
Unzip the downloaded file to a local directory
Install the package from the local directory:
unzip it

## works for me now aftering download the seurat-data-master folder
```{r}
#install.packages("/path/to/local/directory/seurat-data-master", repos = NULL, type="source")
install.packages("seurat-data-master", repos = NULL, type="source")
library(SeuratData)
```

# however, still not able to find lungref data
```{r}
# Load the lungref dataset
data("lungref")
# still not able to find lungref data
```

# not working
```{r}
# C:/Users/pengl7/AppData/Local/R/win-library/4.2/
install.packages("lungref.SeuratData", repos = NULL, type="source") # NOT WORK
```

