---
title: "LP"
output: html_document
date: "2023-09-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This is done on local windows accessing HPC datasets 

```{r}
library(Seurat)
library(SeuratData)
library(patchwork)
library(ggplot2)
library(hdf5r)
```


- the current wd is "Y:/IRF/Cellranger/"


```{r}
combined <- readRDS("rds/epsilon_sct_az_hst.rds")
combined
```

# only test on D3 which has the most infection of Sars

```{r}
Idents(combined) <- combined$orig.ident
d3 <- subset(combined, idents = "D3")
Idents(d3) <- d3$seurat_clusters
d3
```

# using AverageExpression() function to calculate average expression for each cluster
# by default, all the applicable assays inside D3, e.g., RNA, refAssay, integrated, and etc., will do perform the function 
# according to the active idents, and return a matrix for each assay
# the default slot is "data", i.e., log transformed and normalized data
# ideal should be the one from "RNA" assay; do not use "integrated" for expression analysis
# will also generate a table for the prediction.score for each cell types across each cell clusters

# IF only want to chose "RNA" assay, define it inside the funciton of AverageExpression(d3, assay = "RNA", data = "count")
```{r}
# Calculate average expression for each cluster
# here default groups by "ident" on all 6 assays; the returned is a list of matrixes
average_expression <- AverageExpression(d3)
# Save the results to a data frame
for (assay_id in names(average_expression)) {
  cluster_average_expression <- as.data.frame(average_expression[[assay_id]])
  write.csv(cluster_average_expression, file = paste0("d3_epsilon_cluster_", assay_id, "_average_expression.csv"), row.names = TRUE)
}
```

# repeat the above but using slot = "counts")
```{r}
# also generate the pseudo count matrix for DE analysis
average_expression <- AverageExpression(d3, slot = "counts")
for (assay_id in names(average_expression)) {
  cluster_average_expression <- as.data.frame(average_expression[[assay_id]])
  write.csv(cluster_average_expression, file = paste0("d3_epsilon_cluster_", assay_id, "_average_counts.csv"), row.names = TRUE)
}
```

# perform DE on psudo bulk data
# Not work, returned an empty dataframe
```{r}
# get the count data from "RNA" assay
average_expression <- AverageExpression(d3, slot = "counts", assays = "RNA")
average_counts <- as.data.frame(average_expression[["RNA"]])
# Create a Seurat object with pseudobulk data
pseudobulk <- CreateSeuratObject(counts = average_counts)
# Assuming colnames(average_expression) are cluster names
pseudobulk$seurat_clusters <- colnames(average_counts)
# Perform differential expression analysis
de_results <- FindAllMarkers(pseudobulk, test.use = "wilcox")
# fail reason: I have no duplicated for each cluster???
de_results #data frame with 0 columns and 0 rows; Warning: No DE genes identified;
```

```{r}
# Filter results
significant_genes <- de_results[de_results$p_val_adj < 0.05 & abs(de_results$log2fc) > 1, ]

```
